---
title: Estimating the false discovery rate from sample size
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating the false discovery rate from sample size}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

<style type="text/css">

h1.title {
  font-size: 22px
}

</style>

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir="../")
knitr::opts_chunk$set(fig.width=7.5, fig.height=6, fig.path='../figures/', warning=FALSE, message=FALSE,cache=FALSE)

library(phylosamp)
```

This vignette provides an overview of the primary function of the phylosamp package: how to estimate the false discovery rate given a sample size. In the examples provided, we use the default `assumption` argument (multiple transmissions and multiple links, `mtml`), though alternative assumptions can also be specified.

The most basic function of the package is `truediscoveryrate()`, which calculates the probability that an identified link represents a true transmission event:

```{r truediscoveryrate}

truediscoveryrate(eta=0.99, chi=0.95, rho=0.75, M=100, R=1)

```

In other words, given a sample size of 100 infections (representing 75% of the total population), a linkage criteria with a specificity of 99% for identifying infections linked by transmission and a specificity of 95%, fewer than 25% of identified pairs will represent true transmission events. Increasing the specificity to 99.5% has a significant impact on our ability to distinguish linked and unlinked pairs:

```{r truediscoveryrate2}

truediscoveryrate(eta=0.99, chi=0.995, rho=0.75, M=100, R=1)

```

The other core functions are designed to calculate the expected number of true transmission pairs identified in the sample (`true_pairs`) and the total number of linkages one can expect to identify given the sensitivity and specificity of the linkage criteria and a particular sample size and proportion (`exp_links`).

```{r expected_pairs}

true_pairs(eta=0.99, rho=0.75, M=100, R=1)
exp_links(eta=0.99, chi=0.95, rho=0.75, M=100, R=1)

```

##### Important Note

It is important to recognize that $R$ in these functions represents the average $R$ in the sampled population (alternatively denoted $R_\text{pop}$). Because any sampling frame contains a finite number of cases, there will always be more cases than infection events (at minimum, all infectees in a transmission chain plus a single index case), so $R_\text{pop}\leq1$. For outbreaks with a single introduction, $R_\text{pop}$ is approximately equal to 1; sampling frames containing cases from separate introduction events will have lower values of $R_\text{pop}$.